---
title: "nhanes_metals"
author: "Nicole Comfort"
date: "11/26/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install necessary packages
install.packages("tidyverse")
install.packages("NHANES")
install.packages("survey") # Documentation for this package is: https://cran.r-project.org/web/packages/survey/survey.pdf
install.packages("Hmisc")
install.packages("foreign")
install.packages("labelled")

# load libraries
library(tidyverse)
library(NHANES)
library(survey) # This package handles survey weights/designs easily
library(Hmisc) # This is a package (H misc.) for basic desciptive functions.
library(foreign) # This package allows us to open .XPT files
library(labelled)

```

# Background:  
The National Health and Nutrition Examination Survey (NHANES) is a nationally-representative survey of the general, noninstutionalized US population conducted by the National Center for Health Statistics. NHANES is a multi-stage, nationally representative sample in two-year cycles; participants are chosen to represent the entire general US. Certain age groups and racial/ethnic groups are intentionally oversampled to ensure adequate power to analyze these populations.

Participants undergo thorough examination including demographic questionnaire, a 24-hr dietary recall, a clinical examination, and a laboratory examination. All NHANES protocols were approved by the National Center for Health Statistics institutional review board, and all participants gave written informed consent.

For more information on NHANES, view the Centers for Disease Control and Prevention (CDC) website: https://www.cdc.gov/nchs/nhanes/index.htm

### Note:
NHANES does not measure every environmental chemical in every participant. For some environmental chemicals, these measurements are only done on a subset of participant samples and there are special sub-weights that need to be used for analyzing these chemicals. As a general rule, you should use the weight for the smallest of the sampling subset you are analyzing in NHANES. 

Here, we will explore the levels of lead, cadmium, mercury, and manganese measured in blood of males versus females to examine if there is a sex difference in body burden of these chemicals. We will conduct this analysis in one two-year survey cycle (2015/2016).

## Download NHANES Data and Clean


```{r cars}

summary(cars)


```

## Sample Weighting

You can also embed plots, for example:

```{r pressure, echo=FALSE}

# Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

plot(pressure)

```

## Descriptive Statistics

```{r}

```

## Data Visualization

```{r}

```



